Manuscript Figures

### 2021-05-31 src/manuscript_figures.Rmd
add section about immunotherapy response prediction

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = params$workdir)
settings <- yaml::read_yaml(params$settings_file)
FigureFolder <- params$figures_folder
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
library(ggridges)
library(ggcorrplot)
library(ComplexHeatmap)
library(pheatmap)
theme_set(ggpubr::theme_pubclean())
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'

@@

# do unsupervised clustering experiment and compute AMI 
# downsample pan-cancer dataset 
df <- data.frame('id' = rownames(LFs$pca_factors$pancancer.cnv_gex_meth_mut))
df$pr <- surv[match(df$id, bcr_patient_barcode)]$project
ids <- lapply(split(df, df$pr), function(x) {
  if(nrow(x) > 200) {
    sample(x$id, 200)
  }
})
ids <- ids[lengths(ids) == 200]
k <- 5  
# sample cancer types and compute ami for different tools 
sampled <- sapply(simplify = F, 1:50, function(i) {
  unique(unlist(ids[sample(1:length(ids), k)]))
})
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('ids', 'df', 'LFs', 'settings', 'sampled', 'k'))
res <- do.call(rbind, pbapply::pblapply(cl = cl, 1:length(sampled), function(i) {
  require(data.table)
  source(settings$utility_script)
  selected <- sampled[[i]]
  ct <- df[match(selected, df$id),]$pr
  message("selected cancers:",paste0(unique(ct), collapse = ','))
  do.call(rbind, lapply(names(LFs), function(tool) {
    M <- scale(LFs[[tool]]$pancancer.cnv_gex_meth_mut[selected,])
    if(tool == 'maui_factors') {
      M <- scale(M)
    }
    ami <- compute_AMI(M, k, ct)
    return(data.table('i' = i, 'ami' = ami, 'method' = tool, 
                      'pr' = paste0(unique(ct), collapse = ',')))
  }))
}))
parallel::stopCluster(cl)
res$method <- toupper(gsub("_factors", "", res$method))

@@

p3 <- ggplot(ci, aes(x = reorder(project, with_lf))) + 
  geom_segment(aes(x = reorder(project, with_lf), xend = project, 
                   y = base, yend = with_lf)) + 
  geom_point(data = dt, aes(x = project, y = value, color = variable), size = 6) +
  coord_flip() + 
  labs(x = '', y = '') + 
  theme(legend.title = element_blank(), legend.position = 'bottom', 
        legend.background = element_blank(), 
        legend.margin = margin(), 
        legend.direction = 'vertical', legend.box.background = element_blank()) +
  scale_color_brewer(type = 'qual', palette = 6)

@@

# explore MSI related LFs
# tsne plot of top LF related to MSI-H and MSI classes
dt <- get_factor_specific_variables(M, y)
# use top predictive lfs to guide tsne 
top_vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 5)[['variable']])

@@

# scores: geneset scores
# M: survival predictive LF matrix
# maui_clust: best survival prognostic clustering of M 
plot_clusts <- function(M, maui_clust, scores) {
  df <- plot_tsne(M, as.factor(maui_clust[,1]), returnData = T)
  colnames(df)[3] <- 'maui'

  p1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
    geom_point(aes(color = maui)) +
    geom_text_repel(data = get_centroid(df[,1:2], df$maui), 
            aes(x,y,label = text.label), size = 6)
  # plot survival for the worst prognostic group, and the rest
  ms <- get_median_survival_by_group(surv, df[,'maui',drop=F])

@@

  p2 <- plot_survival(surv, 
                      df[df$maui %in% c(best, worst), 
                         'maui', drop = F])[['plot']] +
    labs(y = 'PFI') +
    theme(legend.position = 'left', legend.title = element_blank()) + 
    scale_color_brewer(type = 'qual', palette = 6)

  # heatmap of hallmark scores for best/worst survivors
  selected <- intersect(rownames(df[df$maui %in% c(worst, best),]), rownames(S))
  dt <- get_factor_specific_variables(S[selected,], df[selected, 'maui'])
  # select factors 
  dt[padj < 0.05][order(padj)]
  h1 <- plot_label_specific_vars(S[selected,], 
                                 df[selected, 'maui'], 
                                 padj_threshold = 0.1,
                                as_heatmap = T, cellwidth = 15, 
                                top_vars_per_label = 5, cellheight = 15)
  return(list('p1' = p1, 'p2' = p2, 'h1' = h1))
}

@@

e <- dt[1,]$experiment
M <- prog.clust[[e]]$prog.lf
maui_clust <- prog.clust[[e]]$prog.clust
S <- scores[intersect(rownames(M), rownames(scores)),]
P <- plot_clusts(M, maui_clust, S)

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.semi_supervised.prognostic.clustering.', experiment,'.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', 
       width = 10, height = 12)
Legends[[fig_count]] <- "Log-rank test of prognostic subtypes by MAUI and existing TCGA subtypes. B) tSNE plot of MAUI LFs that are predictive of survival outcomes for TCGA-UCEC. C) Kaplan-Meier curve of the best/worst surviving prognostic MAUI clusters D) Biological pathway scores that distinguish best/worst surviving clusters."
```

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = params$workdir)
settings <- yaml::read_yaml(params$settings_file)
FigureFolder <- params$figures_folder
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
library(ggridges)
library(ggcorrplot)
library(ComplexHeatmap)
library(pheatmap)
library(survminer)
library(survival)
theme_set(ggpubr::theme_pubclean())
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'

@@

# do unsupervised clustering experiment and compute AMI 
# downsample pan-cancer dataset 
df <- data.frame('id' = rownames(LFs$pca_factors$pancancer.cnv_gex_meth_mut))
df$pr <- surv[match(df$id, bcr_patient_barcode)]$project
ids <- lapply(split(df, df$pr), function(x) {
  if(nrow(x) > 200) {
    sample(x$id, 200)
  }
})
ids <- ids[lengths(ids) == 200]
k <- 5
# sample cancer types and compute ami for different tools 
sampled <- sapply(simplify = F, 1:50, function(i) {
  unique(unlist(ids[sample(1:length(ids), k)]))
})
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('ids', 'df', 'LFs', 'settings', 'sampled', 'k'))
res <- do.call(rbind, pbapply::pblapply(cl = cl, 1:length(sampled), function(i) {
  require(data.table)
  source(settings$utility_script)
  selected <- sampled[[i]]
  ct <- df[match(selected, df$id),]$pr
  message("selected cancers:",paste0(unique(ct), collapse = ','))
  do.call(rbind, lapply(names(LFs), function(tool) {
    M <- LFs[[tool]]$pancancer.cnv_gex_meth_mut[selected,]
    if(tool == 'maui_factors') {
      M <- scale(M)
    }
    ami <- compute_AMI(M, k, ct)
    return(data.table('i' = i, 'ami' = ami, 'method' = tool, 
                      'pr' = paste0(unique(ct), collapse = ',')))
  }))
}))
parallel::stopCluster(cl)
res$method <- toupper(gsub("_factors", "", res$method))

@@

p3 <- ggplot(ci, aes(x = reorder(project, with_lf))) + 
  geom_segment(aes(x = reorder(project, with_lf), xend = project, 
                   y = base, yend = with_lf)) + 
  geom_point(data = dt, aes(x = project, y = value, color = variable), size = 2) +
  coord_flip() + 
  labs(x = '', y = '') + 
  theme(legend.title = element_blank(), legend.position = 'bottom', 
        legend.background = element_blank(), 
        legend.margin = margin(), 
        legend.direction = 'vertical', legend.box.background = element_blank()) +
  scale_color_brewer(type = 'qual', palette = 6)

@@

# explore MSI related LFs
# tsne plot of top LF related to MSI-H and MSI classes
dt <- get_factor_specific_variables(M, y)
# use top predictive lfs to guide tsne 
top_vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 10)[['variable']])

@@

# scores: geneset scores
# M: survival predictive LF matrix
# maui_clust: best survival prognostic clustering of M 
plot_clusts <- function(surv, M, maui_clust, scores) {
  df <- plot_tsne(M, as.factor(maui_clust[,1]), returnData = T)
  colnames(df)[3] <- 'maui'

  message("running plot clusts")
  message("p1")
  p1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
    geom_point(aes(color = maui)) +
    geom_text_repel(data = get_centroid(df[,1:2], df$maui), 
            aes(x,y,label = text.label), size = 6)
  # plot survival for the worst prognostic group, and the rest
  ms <- get_median_survival_by_group(surv, df[,'maui',drop=F])

@@

  message("p2")
  p2 <- plot_survival(surv, 
                      df[df$maui %in% c(best, worst), 
                         'maui', drop = F])[['plot']] +
    labs(y = 'PFI') +
    theme(legend.position = 'left', legend.title = element_blank()) + 
    scale_color_brewer(type = 'qual', palette = 6)

  # heatmap of hallmark scores for best/worst survivors
  selected <- intersect(rownames(df[df$maui %in% c(worst, best),]), rownames(S))
  dt <- get_factor_specific_variables(S[selected,], df[selected, 'maui'])
  # select factors 
  dt[padj < 0.05][order(padj)]
  message("h1")

  h1 <- plot_label_specific_vars(S[selected,], 
                                 df[selected, 'maui'], 
                                 padj_threshold = 0.1,
                                as_heatmap = T, cellwidth = 15, 
                                top_vars_per_label = 5, cellheight = 15)
  return(list('p1' = p1, 'p2' = p2, 'h1' = h1))
}

@@

e <- dt[1,]$experiment
M <- prog.clust[[e]]$prog.lf
maui_clust <- prog.clust[[e]]$prog.clust
S <- scores[intersect(rownames(M), rownames(scores)),]
P <- plot_clusts(surv, M, maui_clust, S)

@@

message("Got the figure")

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.semi_supervised.prognostic.clustering.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', 
       width = 10, height = 12)
Legends[[fig_count]] <- "Log-rank test of prognostic subtypes by MAUI and existing TCGA subtypes. B) tSNE plot of MAUI LFs that are predictive of survival outcomes for TCGA-UCEC. C) Kaplan-Meier curve of the best/worst surviving prognostic MAUI clusters D) Biological pathway scores that distinguish best/worst surviving clusters."
```

@@

```{r immunotherapy_response}
# immunotherapy response prediction analysis: 
# TODO: pass paths via args
settings_imtrp <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/immunotherapy_response/Mariathasan_et_al/settings.yaml')
pathways <- readRDS('/data/local/buyar/arcas/subtyping_paper/data/GeneSets/msigdb.pathways.RDS')

LFs_imtrp <- import_LFs(c('maui_factors', 'mofa_factors', 'pca_factors'), settings_imtrp$pipeline_output)
FWs_imtrp <- import_feature_weights(c('maui_factors'), settings_imtrp$pipeline_output)
colData <- data.table::fread(settings_imtrp$clin)
colnames(colData)[1] <- 'bcr_patient_barcode'

dt <- do.call(rbind, lapply(names(LFs_imtrp), function(tool) {
  M <- LFs_imtrp[[tool]]$mariathasan.gex
  # find LFs predictive of response to therapy 
  dt <- get_factor_specific_variables(M, colData$binaryResponse)[order(padj)]
  #dt <- compute_anova(M, colData$binaryResponse)
  dt$tool <- tool
  return(dt)
}))

dt[,.SD[which.min(padj)],by = c('tool', 'ref_cl')]

top_vars <- unique(dt$variable)

# quantify prediction accuracy for finding complete/partial responders
m <- M[,top_vars]
f <- colData$binaryResponse
m <- m[!is.na(f),]
f <- f[!is.na(f)]

# heatmap of top vars 
p1 <- pheatmap::pheatmap(get_basis_matrix(m[,top_vars], f), silent = T)

# survival stratification using the top LF 
require(survminer)
require(survival)
top_lf <- dt[order(padj)][ref_cl == 'CR/PR'][1,]$variable
cp <- find_survival_cutpoint(m[,top_lf,drop = F], colData, event_col = 'censOS', duration_col = 'os')

p2 <- plot_survival(surv = colData, partition.df = data.frame('group' = ifelse(m[,top_lf] > cp, 'high', 'low'), 
                                                        row.names = rownames(m)), 
              duration_col = 'os', event_col = 'censOS')

# top go terms for top features of top LF 
feats <- data.table(W[,top_lf,drop=F], keep.rownames = T)[order(abs(get(top_lf)), decreasing = T)][1:100]
# top pathways for top lf
top_terms <- compute_geneset_enrichment(feats$rn, pathways)[order(pval)][padj < 0.05][1:10]
top_terms$term_wr <- stringr::str_wrap(gsub("_", " ", top_terms$term), width = 32)
p3 <- ggplot(top_terms, aes(x = reorder(term_wr, -pval), y = -log10(padj))) + 
  geom_bar(stat = 'identity') + 
  coord_flip() + labs(x = '') 
colnames(feats)[2] <- 'top_lf'
p4 <- ggplot(feats[1:10,], aes(x = reorder(rn, abs(top_lf)), y = abs(top_lf))) + 
  geom_bar(stat = 'identity') + coord_flip() + labs(x = '', y = 'Feature Importance (absolute)')

# correlation of top LF with tumor mutation burden
# get samples with both response and tmb annotation
cd <- colData #colData[!is.na(`Neoantigen burden per MB`)][!is.na(binaryResponse)]
m2 <- M[cd$bcr_patient_barcode,]
require(ggpubr)
df <- data.frame('response' = cd$binaryResponse, 
                 'TMB' = cd$`Neoantigen burden per MB`, 
                 'LF' = m2[,top_lf])
p5 <- ggpubr::ggscatter(df, x = 'LF', y = 'TMB', 
                        add = 'reg.line', cor.coef = T, cor.coef.coord = c(30, 10), 
                  cor.coef.size = 4) + labs(x = top_lf, y = 'Tumor Mutation Burden')

# immune phenotypes 
ip <- colData$`Immune phenotype`
#ip[ip == 'excluded'] <- NA
m3 <- M[!is.na(ip),]
ip <- ip[!is.na(ip)]

dt3 <- get_factor_specific_variables(m3, ip)[order(pval)][padj < 0.01]
ip_vars <- unique(dt3$variable)
ip_vars <- get_top_by_label(dt3, 'ref_cl', topN = 5)
res_ip <- predict_cluster_labels_glmnet(m3[,unique(dt3$variable)], ip, returnROC = T)
p6 <- plot_roc_list(res_ip)

pheatmap::pheatmap(get_basis_matrix(m3[,unique(ip_vars$variable)], ip), silent = T)

Fig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2$gtable, p4, p3$plot, 
                                             labels = 'AUTO'), 
                          cowplot::plot_grid(p5, p6, p7, nrow = 1, 
                                             labels = c('E', 'F', 'G')), 
                          nrow = 2, rel_heights = c(2, 1)) 
fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.immunotherapy_response.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', 
       width = 10, height = 14)
Legends[[fig_count]] <- "Predicting responders to anti-PD-L1 treatment in a cohort of Metastatic Urothelial
Cancers (N=348) using MAUI. A) AUC curve for predicting Complete/Partial Responders (CR/PR) to the anti-PD-L1
treatment using an Elastic Net Model on a subset of MAUI latent factors individually predictive of treatment response groups. B) Average activation score of top latent factors predictive of treatment response groups. C) Top pathways enriched among top 100 genes that contribute to the top latent factor (LF44). D) Survival stratification of the 
cohort based activation score of LF44 (high = LF44 > 1.4, low = LF44 < 1.4). E) Top 10 genes contributing to LF44
F) Correlation between LF44 activation score and the tumor (neo-antigen) mutation burden"
```
